---
title: "Sonja Next Word: Speed and Quality Check"
subtitle: "A faster response and a controlled compression comparison"
author: "Author: Sonja Sahebzad"
date: "2026-09-25"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
    css: ../../../milestone-report.css
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE)
d <- read.csv("development_comparison.csv")
s <- jsonlite::fromJSON("selection.json")
stopifnot(s$selected=="Current",!s$final_test_used)
base <- d[d$name=="Current",]
best <- max(d$top3)
pct <- function(x) sprintf("%.1f%%",100*x)
labels <- c(Current="Current",More_contexts="More contexts",More_followers="More followers",Larger_vocabulary="Larger vocabulary")
```

<div class="summary-box">
**Management summary.** Automatic suggestions now wait **150 ms instead of 400 ms** after a completed word and a pause. This removes 250 ms of intentional delay without changing any prediction. Three changes to model compression were tested separately on the same 600 development cases. The best result was **`r best`/600**, compared with **`r base$top3`/600** for the current model. No challenger met the predefined replacement rule. The deployed model and its independent **28.6% top-three accuracy** remain unchanged.
</div>

*Later update: an [exact indexed implementation](https://sonja242.github.io/data-science-capstone-swiftkey/context-speed-check.html) reduced local computation further to 0.58 ms at the median, with unchanged predictions. The compression comparison below remains the record of this earlier experiment.*

# What became faster

At the time of this experiment, median local computation was **1.9 ms**. Automatic waiting, network transport, server load and browser rendering add to the time the user experiences. The delay setting fell by **62.5%**; this is not a claim that total response time fell by 62.5%. Manual prediction bypasses this typing delay. The app still waits for a word boundary, preventing automatic predictions from unfinished words.

Server tests verified that no result appears at 100 ms and a result is available after 160 ms of simulated time. Tests also covered rapid edits, clearing while a timer is pending, newlines, manual mode and input limits. All prediction words and scores remained identical on 1,500 existing development/test examples. These checks establish regression equivalence, not a new accuracy estimate or a real-network latency benchmark.

# Controlled model comparison

The 600 development examples contain 200 cases from each source. Each challenger changes **one** compression setting while retaining the same trained weights, scoring method and maximum four-word context. More contexts lowers minimum support from five to two. More followers retains up to sixteen continuations instead of eight. Larger vocabulary retains 100,000 words instead of 50,000.

Before comparison, the rule required at least **six additional top-three successes**, no loss in top-one successes, and limits of 256 MiB model memory and 50 MiB compressed size. This study increased the earlier 200 MiB memory limit before comparison to allow a larger candidate, while retaining hosting headroom. The complete protocol is in `protocol.json`. Timing is one sequential local pass; small timing differences do not establish a reliable speed ranking.

```{r comparison}
knitr::kable(data.frame(Variant=unname(labels[d$name]),
  `First correct`=sprintf("%d/600 (%s)",d$top1,pct(d$top1/600)),
  `Within three`=sprintf("%d/600 (%s)",d$top3,pct(d$top3/600)),
  `Memory (MiB)`=round(d$object_mib,1),
  `Median computation (ms)`=round(d$median_ms,2),check.names=FALSE),
  caption="Table 1. Same 600 development cases for every configuration. Timing excludes network and typing delay.")
```

```{r comparison-chart}
plotly::plot_ly(x=unname(labels[d$name]),y=100*d$top3/600,type="bar",
  marker=list(color=c("#123f68","#75aacf","#75aacf","#75aacf")),
  text=sprintf("%s<br>%d/600: %.1f%%<br>Model memory: %.1f MiB",unname(labels[d$name]),d$top3,100*d$top3/600,d$object_mib),
  hoverinfo="text") |>
  plotly::layout(yaxis=list(title="Correct within three (%)",range=c(0,35)),
    xaxis=list(title=""),font=list(family="Calibri, Segoe UI, sans-serif"))
```

*Figure 1. Development accuracy. Hover over a bar to see its count and model memory.*

# Discussion

The largest observed gain was **`r best-base$top3` additional correct words out of 600**, or **`r sprintf("%.2f",100*(best-base$top3)/600)` percentage points**. All three challengers require more memory. None satisfies the predefined minimum improvement. The evidence therefore does not justify replacing the tested app model with one of these variants.

These are reused development data, so their percentages should not be presented as independent test performance. They also cannot be compared directly with the app's 28.6% on a different 900-case sample. Its existing approximate 95% interval remains **25.7% to 31.6%**. This interval describes sampling uncertainty in aggregate accuracy, not confidence in an individual suggestion. No calibrated confidence percentage is displayed.

The test checks an exact match with the original next word. A different plausible continuation is counted as incorrect. This explains what the measure captures; it does not establish that alternative suggestions are useful. All candidates retain the same short context and trained model, so this experiment does not test stronger sentence-level understanding.

# Conclusion

The app now responds sooner after completed words, with unchanged predictions. Loosening compression alone did not meet the required quality improvement. A stronger follow-up should compare a model that uses more sentence context under an explicit latency and memory budget, then assess it on untouched cases. A user study should separately measure whether people accept suggestions and save typing effort. A high quiz score or a different model's accuracy cannot substitute for this evidence.

# Reproducibility and references

This report was knitted from R Markdown. `compare_quality.R` records the protocol before comparison, changes one setting at a time and writes case-level outcomes without publishing text. Seeds 20260927 and 20260928 reserved 900 fresh test cases after excluding previous evaluations and checking train/validation separation. **These cases remain unscored**, because no development challenger qualified. Their hashes are in `reserved_hashes.csv` and must be excluded from other experiments until this reservation is released. `selection.json` records the decision. Models and full text stay in the private, ignored `data/final_product_quality` folder. `session-info.txt` records package versions.

1. [Official Coursera SwiftKey corpus](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip).
2. Chen and Goodman (1996). [An Empirical Study of Smoothing Techniques for Language Modeling](https://aclanthology.org/P96-1041/).
3. Posit. [Shiny debounce documentation](https://shiny.posit.co/r/reference/shiny/latest/debounce.html) and [shinyapps.io application resources](https://docs.posit.co/shinyapps.io/guide/applications/).
