---
title: "Practice: File and Text Operations in R"
author: "Author: Sonja Sahebzad"
date: "2026-09-19"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

<style>
body { font-family: "Segoe UI", Calibri, sans-serif; color: #111111; }
h1, h2, h3 { color: #174a7e; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This notebook teaches the R operations needed to inspect text files. It uses a small practice corpus and does not contain answers to the monitored Coursera quiz.

# Practice text

```{r practice-data}
practice_lines <- c(
  "Data products turn analysis into useful tools.",
  "A clear model can support a decision.",
  "Good data needs careful cleaning.",
  "A model is useful when its errors are understood.",
  "This exact practice sentence appears once."
)

practice_lines
```

# Inspect a file size

R reports file sizes in bytes. Dividing by `1024^2` converts bytes to mebibytes, which are commonly displayed as MB.

```{r file-size}
practice_file <- tempfile(fileext = ".txt")
writeLines(practice_lines, practice_file)

size_bytes <- file.info(practice_file)$size
size_mb <- size_bytes / 1024^2

data.frame(size_bytes = size_bytes, size_mb = size_mb)
```

# Count lines

```{r count-lines}
length(practice_lines)
```

For a large file, process chunks so the complete corpus is not retained in memory.

```{r chunked-line-count}
count_lines_chunked <- function(path, chunk_size = 10000L) {
  connection <- file(path, open = "r", encoding = "UTF-8")
  on.exit(close(connection))

  total <- 0L
  repeat {
    chunk <- readLines(connection, n = chunk_size, warn = FALSE, skipNul = TRUE)
    if (length(chunk) == 0L) break
    total <- total + length(chunk)
  }
  total
}
```

# Find the longest line

```{r longest-line}
line_lengths <- nchar(practice_lines, type = "chars")
longest_index <- which.max(line_lengths)

data.frame(
  line_number = longest_index,
  length = line_lengths[longest_index],
  text = practice_lines[longest_index]
)
```

# Find a lowercase word

Word boundaries ensure that a search term is treated as a complete word. The search below is case-sensitive.

```{r find-word}
contains_data <- grepl("\\bdata\\b", practice_lines, perl = TRUE)
practice_lines[contains_data]
sum(contains_data)
```

# Compare two search counts

```{r compare-counts}
model_count <- sum(grepl("\\bmodel\\b", practice_lines, perl = TRUE))
error_count <- sum(grepl("\\berrors?\\b", practice_lines, perl = TRUE))

if (error_count > 0L) model_count / error_count else NA_real_
```

# Search for an exact complete line

```{r exact-match}
target_line <- "This exact practice sentence appears once."
sum(practice_lines == target_line)
```

# Your independent practice

Use the same ideas on a text file of your choice:

1. Inspect its size and state the unit used.
2. Count its lines without retaining the complete file.
3. Find the longest line and identify its position.
4. Count lines containing two different lowercase words.
5. Calculate a ratio only after checking that the denominator is nonzero.
6. Search for one complete line using exact equality.

Explain each result in your own words before starting the monitored quiz.


# Your turn: count the official Twitter file

Run this chunk yourself in RStudio. It checks that the official file exists and then counts its lines without keeping the complete file in memory.

```{r official-twitter-line-count, eval=FALSE}
twitter_file <- file.path("data", "final", "en_US", "en_US.twitter.txt")
file.exists(twitter_file)
count_lines_chunked(twitter_file)
```

# Six-step self-study worksheet

Use this worksheet to practise each required operation. Replace only the quoted placeholders with a file path, search word, or sentence chosen for your own work. Run each chunk yourself and explain the output in your own words.

## Task 1 — Inspect a file size

```{r worksheet-file-size, eval=TRUE}
file_path <- practice_file
size_bytes <- file.info(file_path)$size
size_mib <- size_bytes / 1024^2
data.frame(size_bytes, size_mib)
```

## Task 2 — Count lines

```{r worksheet-line-count, eval=TRUE}
file_path <- practice_file
count_lines_chunked(file_path)
```


## Task 3 — Find the longest line

```{r worksheet-longest-line, eval=TRUE}
practice_files <- c(
  source_a = tempfile(fileext = ".txt"),
  source_b = tempfile(fileext = ".txt"),
  source_c = tempfile(fileext = ".txt")
)

writeLines(c("Short line.", "A somewhat longer practice line."), practice_files[1])
writeLines(c("Tiny.", "This neutral example is the longest practice line in the three files."), practice_files[2])
writeLines(c("Another line.", "A medium practice sentence."), practice_files[3])

longest_per_file <- vapply(practice_files, function(path) {
  lines <- readLines(path, warn = FALSE, skipNul = TRUE)
  max(nchar(lines, type = "chars"))
}, integer(1))

data.frame(file = names(longest_per_file), longest_line = longest_per_file)
```

## Task 4 — Compare two lowercase word counts

```{r worksheet-word-ratio, eval=TRUE}
text_lines <- practice_lines
word_a <- "data"
word_b <- "model"
count_a <- sum(grepl(paste0("\\b", word_a, "\\b"), text_lines, perl = TRUE))
count_b <- sum(grepl(paste0("\\b", word_b, "\\b"), text_lines, perl = TRUE))
if (count_b > 0L) count_a / count_b else NA_real_
```

## Task 5 — Find a line containing one word

```{r worksheet-find-word, eval=TRUE}
search_word <- "cleaning"
matches <- grepl(paste0("\\b", search_word, "\\b"), text_lines, perl = TRUE)
text_lines[matches]
```

## Task 6 — Count an exact complete line

```{r worksheet-exact-line, eval=TRUE}
target_line <- "This exact practice sentence appears once."
sum(text_lines == target_line)
```
