---
title: "Independent Text-File Verification Toolkit"
author: "Sonja Sahebzad"
date: "2026-09-19"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source(file.path("R", "text_query_utils.R"))
```

# Purpose

This toolkit demonstrates general, memory-safe methods for independently checking large text files. It contains no assessment-specific search terms or answers. Replace the example inputs yourself when practising with another text file.

# Create a neutral practice file

```{r practice-file}
practice_lines <- c(
  "A clear model supports a decision.",
  "Good data requires careful cleaning.",
  "A model is useful when its errors are understood.",
  "This exact practice sentence appears once.",
  "Reliable data supports a reliable model."
)

practice_path <- tempfile(fileext = ".txt")
writeLines(practice_lines, practice_path)
```

# Maximum line length

```{r maximum-length}
maximum_line_length(practice_path)
```

# Count lines containing complete words

```{r word-counts}
first_count <- count_lines_with_word(
  practice_path,
  word = "data",
  case_sensitive = TRUE
)

second_count <- count_lines_with_word(
  practice_path,
  word = "model",
  case_sensitive = TRUE
)

data.frame(first_count, second_count)
```

# Calculate a ratio safely

```{r safe-ratio}
safe_count_ratio(first_count, second_count)
```

# Return matching lines

```{r matching-lines}
find_lines_with_word(
  practice_path,
  word = "errors",
  case_sensitive = TRUE
)
```

# Count exact complete lines

```{r exact-line-count}
count_exact_lines(
  practice_path,
  target = "This exact practice sentence appears once."
)
```

# Independent workflow

1. Select the required source file yourself.
2. Supply the requested search word or complete line yourself.
3. Run the appropriate function in the RStudio console.
4. Record the printed result in your own notes.
5. Check whether case sensitivity, complete-word matching and exact equality match the wording of the task.

The functions process files in chunks, so they can be used without loading an entire large corpus into memory.
