---
title: "SwiftKey Capstone Milestone Report"
subtitle: "Exploratory Analysis and Prediction Product Plan"
author: "Author: Sonja Sahebzad"
date: "2026-09-19"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    css: milestone-report.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%"
)
library(ggplot2)
source(file.path("R", "quiz1_verification.R"))
source(file.path("R", "corpus_pipeline.R"))
source(file.path("R", "milestone_analysis.R"))

blue <- "#1f73a8"
dark_blue <- "#123f68"
light_blue <- "#80b6d6"
```

<div class="summary-box">
<strong>Management summary.</strong> The official English SwiftKey corpus contains three large text sources: blogs, news and Twitter. Together they provide more than four million lines of natural language. The analysis confirms that the files are usable, but also shows that their writing styles differ and that rare word sequences are common. The planned product will therefore use a compact back-off n-gram model, validate predictions on unseen text, filter unsuitable suggestions and return three next-word options quickly in a Shiny app.
</div>

# Data and scope

The analysis uses the unmodified English files supplied through the Coursera Capstone download. The files are processed in chunks, so the complete corpus does not have to be held in memory. A deterministic sample is retained for exploratory plots and language-pattern summaries.

```{r build-artifacts}
paths <- official_corpus_paths()
cache_file <- file.path("data", "derived", "milestone_artifacts.rds")

if (file.exists(cache_file)) {
  artifacts <- readRDS(cache_file)
} else {
  dir.create(dirname(cache_file), recursive = TRUE, showWarnings = FALSE)
  artifacts <- build_milestone_artifacts(paths)
  saveRDS(artifacts, cache_file)
}

corpus_summary <- artifacts$summary
line_sample <- artifacts$sample
```

```{r summary-table}
display_summary <- transform(
  corpus_summary,
  size_mib = round(size_mib, 1),
  lines = format(round(lines), big.mark = ",", scientific = FALSE),
  words = format(round(words), big.mark = ",", scientific = FALSE),
  words_per_line = round(words_per_line, 1),
  sampled_lines = format(sampled_lines, big.mark = ",", scientific = FALSE)
)
names(display_summary) <- c(
  "Source", "File", "Size (MiB)", "Lines", "Word tokens",
  "Words per line", "Sampled lines"
)
knitr::kable(display_summary, caption = "Table 1. Basic summaries of the three official English files", align = c("l", "l", "r", "r", "r", "r", "r"))
```

The three sources are similar in file size but not in structure. Twitter contains many more, shorter records; blogs contain fewer, longer records. This matters because a model trained without source balancing could learn platform-specific writing habits rather than general English patterns.

# Scale of the corpus

```{r corpus-scale-plot}
plot_scale <- rbind(
  data.frame(source = corpus_summary$source, measure = "Lines", millions = corpus_summary$lines / 1e6),
  data.frame(source = corpus_summary$source, measure = "Word tokens", millions = corpus_summary$words / 1e6)
)

ggplot(plot_scale, aes(source, millions, fill = measure)) +
  geom_col(position = "dodge", width = 0.68) +
  geom_text(aes(label = sprintf("%.1f", millions)), position = position_dodge(width = 0.68), vjust = -0.35, size = 3.8) +
  scale_fill_manual(values = c("Lines" = light_blue, "Word tokens" = dark_blue)) +
  labs(title = "Corpus scale by source", subtitle = "Counts shown in millions", x = NULL, y = "Millions", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
```

# How line length differs

```{r line-length-plot}
line_limit <- unname(quantile(line_sample$characters, 0.99, na.rm = TRUE))
line_plot_data <- line_sample[line_sample$characters <= line_limit, ]

ggplot(line_plot_data, aes(characters, fill = source)) +
  geom_histogram(bins = 35, alpha = 0.85, color = "white") +
  facet_wrap(~ source, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c(blogs = "#275f8e", news = "#568fb5", twitter = "#8ebbd2")) +
  labs(
    title = "Sampled line-length distributions",
    subtitle = "The upper 1% is omitted so the typical range remains readable",
    x = "Characters per line",
    y = "Sampled lines"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

Twitter is concentrated in short lines, while blogs and news include longer passages. The prediction engine should be tested on held-out examples from every source so that one style does not dominate the performance estimate.

# Frequent language patterns

```{r top-terms-table-plot}
top_terms <- sample_top_terms(line_sample$text, n = 15L)

ggplot(top_terms, aes(frequency, term)) +
  geom_col(fill = blue, width = 0.72) +
  labs(
    title = "Most frequent informative terms in the sample",
    subtitle = "Common function words were removed for this exploratory view",
    x = "Occurrences",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

The distribution is strongly uneven: a small group of terms occurs often, while many terms and phrases occur only once. This is a standard sparse-language problem. It supports pruning very rare sequences and using shorter contexts whenever a longer phrase has not been seen often enough.

```{r ngram-profile}
ngram_summary <- ngram_profile(line_sample$text, maximum_n = 3L)
ngram_display <- transform(
  ngram_summary,
  observed_tokens = format(observed_tokens, big.mark = ",", scientific = FALSE),
  unique_sequences = format(unique_sequences, big.mark = ",", scientific = FALSE),
  singleton_share_percent = sprintf("%.1f%%", singleton_share_percent)
)
names(ngram_display) <- c("Model", "Observed sequences", "Unique sequences", "Seen once")
knitr::kable(ngram_display, caption = "Table 2. Sample vocabulary growth as context length increases", align = c("l", "r", "r", "r"))
```

Longer sequences carry more context, but they are also much sparser. A prediction system must therefore balance specificity with coverage.

# Prediction algorithm plan

The first production candidate will be a **frequency-based back-off n-gram model**:

1. Normalize case, URLs, punctuation and whitespace while preserving meaningful apostrophes.
2. Build unigram, bigram, trigram and four-gram counts from the training portion.
3. Use the longest available context first and back off to shorter contexts when evidence is limited.
4. Apply discounting and prune extremely rare sequences to control memory and response time.
5. Exclude profanity and return the three highest-ranked next-word candidates.

The model will be compared with simpler baselines. A held-out validation set will measure top-1 accuracy, top-3 accuracy, mean reciprocal rank, coverage, response time and model size. The final choice will balance predictive quality with the speed required for an interactive product.

# Shiny product plan

The app will accept a short English phrase and immediately display three suggested next words. It will also show a plain-language explanation of the context used, allow the user to append a suggestion to the phrase and provide example prompts. The interface will be mobile-friendly because predictive typing is primarily a mobile use case.

The first release will focus on reliable prediction and fast response. Diagnostic information such as model coverage and fallback level can be available in an optional panel without distracting the main user flow.

# Discussion

The complete corpus contains 4,269,642 lines and 102,404,647 word tokens (Table 1), but the three sources contribute different types of evidence. Twitter provides 2,360,148 lines, or 55.3% of all records, while contributing 30,096,690 words, or 29.4% of all tokens. Its average of 12.8 words per line is substantially below the 34.4 words in news and 41.8 words in blogs. The line-length distributions show the same pattern: in the deterministic sample, the median line contains 64 characters for Twitter, compared with 167 for blogs and 185.5 for news. A model based only on pooled frequencies could therefore overrepresent short, conversational Twitter contexts. Source-aware sampling and separate validation results by source will make that risk visible.

Table 2 quantifies the sparsity problem that drives the modelling strategy. In the sample, 54.2% of unique unigrams occur once; this rises to 84.3% for bigrams and 95.9% for trigrams. Longer word sequences contain more context but have much poorer coverage. These results support a back-off approach: use the longest context with sufficient evidence, then fall back to shorter sequences when the input is rare or unseen. Pruning very rare sequences should also reduce model size with limited loss of reliable information. This interpretation is a modelling hypothesis rather than a performance claim; held-out accuracy and coverage must establish whether the proposed choices work.

The analysis has several limitations. File-level counts use the complete official corpus, whereas the distribution and n-gram analyses use a deterministic sample of 6,313 lines to control memory and ensure reproducibility. The corpus also mixes sources, dates and writing styles and contains spelling variation, foreign-language fragments and unsuitable terms. Consequently, future performance estimates will describe held-out data from this corpus and will not be generalized to all English writing without external validation.

## Risks and next steps

The next stage will compare unigram, bigram, trigram and four-gram candidates with alternative pruning and back-off rules. A fixed train-validation split will be used to report top-1 accuracy, top-3 accuracy, mean reciprocal rank, coverage, response time and model size, both overall and by source. Profanity filtering and tests with phrases from outside the training sample will address product risk. The selected model must improve predictive usefulness enough to justify any added memory and latency.

# Conclusion

The exploratory analysis confirms that the official SwiftKey corpus was downloaded and processed successfully: it contains 4.27 million lines and 102.40 million word tokens. Two findings are decisive for the product design. First, Twitter accounts for most records but has much shorter text than blogs and news. Second, 95.9% of sampled trigram types occur only once, showing that a fixed long-context model would have poor coverage. The evidence therefore supports testing a compact, source-aware back-off n-gram model with pruning and three next-word suggestions. Predictive performance has not yet been established; the next milestone is to validate candidate models before implementing the final Shiny app.

# Reproducibility

This report was knitted from R Markdown. Essential code and its results are retained in the document and can be revealed with the expandable **Code** buttons. The analysis uses the official corpus files stored under `data/final/en_US/`; cached derived summaries allow the report to be rebuilt without silently changing the underlying inputs.

```{r session-info}
cat("R version:", R.version.string, "\n")
cat("Analysis date: 2026-09-19\n")
```

# References

- Coursera and SwiftKey. *Data Science Capstone: Coursera-SwiftKey Dataset*. Official course dataset: <https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip> (accessed September 19, 2026).
- HC Corpora. *About the Corpora*. Archived description of collection, language filtering and anonymisation: <https://web.archive.org/web/20160930083655/http://www.corpora.heliohost.org/aboutcorpus.html> (accessed September 19, 2026).
- Jurafsky, D., and Martin, J. H. *Speech and Language Processing*. Online manuscript: <https://web.stanford.edu/~jurafsky/slp3/> (accessed September 19, 2026).
- Katz, S. M. (1987). Estimation of probabilities from sparse data for the language model component of a speech recognizer. *IEEE Transactions on Acoustics, Speech, and Signal Processing*, 35(3), 400-401. <https://doi.org/10.1109/TASSP.1987.1165125>.
