---
title: "SwiftKey Capstone Milestone Report"
subtitle: "Exploratory Analysis and Prediction Product Plan"
author: "Author: Sonja Sahebzad"
date: "2026-09-19"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    css: milestone-report.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%"
)
library(ggplot2)
source(file.path("R", "quiz1_verification.R"))
source(file.path("R", "corpus_pipeline.R"))
source(file.path("R", "milestone_analysis.R"))

blue <- "#1f73a8"
dark_blue <- "#123f68"
light_blue <- "#80b6d6"
```

<div class="summary-box">
<strong>Management summary.</strong> The official English SwiftKey corpus contains three large text sources: blogs, news and Twitter. Together they provide more than four million lines of natural language. The analysis confirms that the files are usable, but also shows that their writing styles differ and that rare word sequences are common. The planned product will therefore use a compact back-off n-gram model, validate predictions on unseen text, filter unsuitable suggestions and return three next-word options quickly in a Shiny app.
</div>

# Data and scope

The analysis uses the unmodified English files supplied through the Coursera Capstone download. The files are processed in chunks, so the complete corpus does not have to be held in memory. A deterministic sample is retained for exploratory plots and language-pattern summaries.

```{r build-artifacts}
paths <- official_corpus_paths()
cache_file <- file.path("data", "derived", "milestone_artifacts.rds")

if (file.exists(cache_file)) {
  artifacts <- readRDS(cache_file)
} else {
  dir.create(dirname(cache_file), recursive = TRUE, showWarnings = FALSE)
  artifacts <- build_milestone_artifacts(paths)
  saveRDS(artifacts, cache_file)
}

corpus_summary <- artifacts$summary
line_sample <- artifacts$sample
```

```{r summary-table}
display_summary <- transform(
  corpus_summary,
  size_mib = round(size_mib, 1),
  lines = format(round(lines), big.mark = ",", scientific = FALSE),
  words = format(round(words), big.mark = ",", scientific = FALSE),
  words_per_line = round(words_per_line, 1),
  sampled_lines = format(sampled_lines, big.mark = ",", scientific = FALSE)
)
names(display_summary) <- c(
  "Source", "File", "Size (MiB)", "Lines", "Word tokens",
  "Words per line", "Sampled lines"
)
knitr::kable(display_summary, caption = "Table 1. Basic summaries of the three official English files", align = c("l", "l", "r", "r", "r", "r", "r"))
```

The three sources are similar in file size but not in structure. Twitter contains many more, shorter records; blogs contain fewer, longer records. This matters because a model trained without source balancing could learn platform-specific writing habits rather than general English patterns.

# Scale of the corpus

```{r corpus-scale-plot}
plot_scale <- rbind(
  data.frame(source = corpus_summary$source, measure = "Lines", millions = corpus_summary$lines / 1e6),
  data.frame(source = corpus_summary$source, measure = "Word tokens", millions = corpus_summary$words / 1e6)
)

ggplot(plot_scale, aes(source, millions, fill = measure)) +
  geom_col(position = "dodge", width = 0.68) +
  geom_text(aes(label = sprintf("%.1f", millions)), position = position_dodge(width = 0.68), vjust = -0.35, size = 3.8) +
  scale_fill_manual(values = c("Lines" = light_blue, "Word tokens" = dark_blue)) +
  labs(title = "Corpus scale by source", subtitle = "Counts shown in millions", x = NULL, y = "Millions", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
```

# How line length differs

```{r line-length-plot}
line_limit <- unname(quantile(line_sample$characters, 0.99, na.rm = TRUE))
line_plot_data <- line_sample[line_sample$characters <= line_limit, ]

ggplot(line_plot_data, aes(characters, fill = source)) +
  geom_histogram(bins = 35, alpha = 0.85, color = "white") +
  facet_wrap(~ source, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c(blogs = "#275f8e", news = "#568fb5", twitter = "#8ebbd2")) +
  labs(
    title = "Sampled line-length distributions",
    subtitle = "The upper 1% is omitted so the typical range remains readable",
    x = "Characters per line",
    y = "Sampled lines"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

Twitter is concentrated in short lines, while blogs and news include longer passages. The prediction engine should be tested on held-out examples from every source so that one style does not dominate the performance estimate.

# Frequent language patterns

```{r top-terms-table-plot}
top_terms <- sample_top_terms(line_sample$text, n = 15L)

ggplot(top_terms, aes(frequency, term)) +
  geom_col(fill = blue, width = 0.72) +
  labs(
    title = "Most frequent informative terms in the sample",
    subtitle = "Common function words were removed for this exploratory view",
    x = "Occurrences",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

The distribution is strongly uneven: a small group of terms occurs often, while many terms and phrases occur only once. This is a standard sparse-language problem. It supports pruning very rare sequences and using shorter contexts whenever a longer phrase has not been seen often enough.

```{r ngram-profile}
ngram_summary <- ngram_profile(line_sample$text, maximum_n = 3L)
ngram_display <- transform(
  ngram_summary,
  observed_tokens = format(observed_tokens, big.mark = ",", scientific = FALSE),
  unique_sequences = format(unique_sequences, big.mark = ",", scientific = FALSE),
  singleton_share_percent = sprintf("%.1f%%", singleton_share_percent)
)
names(ngram_display) <- c("Model", "Observed sequences", "Unique sequences", "Seen once")
knitr::kable(ngram_display, caption = "Table 2. Sample vocabulary growth as context length increases", align = c("l", "r", "r", "r"))
```

Longer sequences carry more context, but they are also much sparser. A prediction system must therefore balance specificity with coverage.

# Prediction algorithm plan

The first production candidate will be a **frequency-based back-off n-gram model**:

1. Normalize case, URLs, punctuation and whitespace while preserving meaningful apostrophes.
2. Build unigram, bigram, trigram and four-gram counts from the training portion.
3. Use the longest available context first and back off to shorter contexts when evidence is limited.
4. Apply discounting and prune extremely rare sequences to control memory and response time.
5. Exclude profanity and return the three highest-ranked next-word candidates.

The model will be compared with simpler baselines. A held-out validation set will measure top-1 accuracy, top-3 accuracy, mean reciprocal rank, coverage, response time and model size. The final choice will balance predictive quality with the speed required for an interactive product.

# Shiny product plan

The app will accept a short English phrase and immediately display three suggested next words. It will also show a plain-language explanation of the context used, allow the user to append a suggestion to the phrase and provide example prompts. The interface will be mobile-friendly because predictive typing is primarily a mobile use case.

The first release will focus on reliable prediction and fast response. Diagnostic information such as model coverage and fallback level can be available in an optional panel without distracting the main user flow.

# Discussion

The exploratory results show that source type is a material modelling choice. Twitter contributes the largest number of lines, but those lines are much shorter than blog and news entries. Training only on pooled counts could therefore favour Twitter-style language. The modelling pipeline will preserve the source labels, use balanced training samples where appropriate and report validation results both overall and by source.

The n-gram profile also confirms a trade-off between context and coverage. Longer phrases can produce more relevant predictions, but many occur only once. A back-off model is therefore preferable to relying on a single n-gram order: it can use detailed context when evidence is strong and shorter context when a phrase is rare or unseen. Candidate models will include unigram, bigram, trigram and four-gram variants with alternative pruning thresholds. The final selection will test whether the main performance conclusion changes across these reasonable choices.

The analysis has limitations. Exploratory language-pattern plots use a deterministic sample to keep processing reproducible and memory-conscious, while the file-level counts use the complete corpus. The corpus also mixes dates and writing styles and includes spelling variation, foreign-language fragments and unsuitable terms. Performance on held-out corpus data will therefore not be presented as performance on all English writing.

## Risks and next steps

Before release, I will complete the train-validation split, compare back-off and pruning rules, validate the profanity filter, measure response time and model size, and test the app with phrases from outside the training sample. These checks will determine whether the improvement from additional context justifies the added complexity and memory use.

# Conclusion

The official English SwiftKey files were downloaded, loaded and summarized successfully. The corpus contains more than four million lines and shows clear differences between blogs, news and Twitter, together with the expected sparsity of longer word sequences. These findings support a compact, source-aware back-off n-gram model that returns three next-word suggestions. The project is ready to proceed from exploratory analysis to model validation and development of the Shiny prediction app.

# Reproducibility

This report was knitted from R Markdown. Essential code and its results are retained in the document and can be revealed with the expandable **Code** buttons. The analysis uses the official corpus files stored under `data/final/en_US/`; cached derived summaries allow the report to be rebuilt without silently changing the underlying inputs.

```{r session-info}
cat("R version:", R.version.string, "\n")
cat("Analysis date: 2026-09-19\n")
```

# References

- Coursera and SwiftKey. *Data Science Capstone: Coursera-SwiftKey Dataset*. Official course dataset: <https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip> (accessed September 19, 2026).
- HC Corpora. *About the Corpora*. Archived description of collection, language filtering and anonymisation: <https://web.archive.org/web/20160930083655/http://www.corpora.heliohost.org/aboutcorpus.html> (accessed September 19, 2026).
- Jurafsky, D., and Martin, J. H. *Speech and Language Processing*. Online manuscript: <https://web.stanford.edu/~jurafsky/slp3/> (accessed September 19, 2026).
- Katz, S. M. (1987). Estimation of probabilities from sparse data for the language model component of a speech recognizer. *IEEE Transactions on Acoustics, Speech, and Signal Processing*, 35(3), 400-401. <https://doi.org/10.1109/TASSP.1987.1165125>.
